Building Markov State Model for a Periodically Driven Non-Equilibrium System

### 2014-08-28 writedown/period.noneq.msm/paper.tex
add paper.tex, merging with C.S.

We denote the phase space by $\Omega$, and the variable by $\vect
x$. The time-dependent probability distribution on the phase space is
denoted by $\rho(\vect x, t)$. The dynamics of the system is governed
by
\begin{align}
  \label{eqn:tmp1}
  \frac{\partial \rho(\vect x, t)}{\partial t} = \ml(t) \rho(\vect x,t),
\end{align}
where the generator $\ml(t)$ is time-dependent and assumed to be
periodic, i.e.~$\ml(t) = \ml(t+T)$, with period denoted by $T$.  Now
we discretize the phase space $\Omega$ into finite number of disjoint
sets $\{ S_1, \cdots, S_n\}$, which satisfy $\Omega = \cup_i S_i$,
$S_j\cap S_j = \emptyset,\ \forall i\neq j$.
Now the original dynamics of the system can be discretized to a time-dependent
Markov jump process:
\begin{align}
  \label{eqn:tmp2}
  \frac{d\vect p(t)}{dt} = \vect M(t)\cdot \vect p(t)
\end{align}
where
\begin{align}
  \vect p(t) = (p_1(t), \cdots, p_n(t))^T, \quad \textrm{with}\ p_i(t) = \int_{S_i} \rho(\vect x,t)d\vect x,
\end{align}
and the generator is approximated by finite difference scheme:
\begin{align}
  \label{eqn:tmp4}
  M_{ij}(t) \approx \frac{1}{\tau} \,[\, \mbp (\vect x_{t+\tau} \in S_j \vert \vect x_{t} \in S_i) - \delta_{ij} \,]
\end{align}
From Eqn.~\eqref{eqn:tmp1} to Eqn.~\eqref{eqn:tmp2} involves several
approximations:
\begin{enumerate}
\item Eqn.~\eqref{eqn:tmp2} assumes Markovianity, which is not
  necessarily true for a reduce dynamics $\vect p(t)$, see e.g.~\cite{prinz2011markov}.
\item Numerical error from the finite difference introduced in
  Eqn.~\eqref{eqn:tmp4}. The lag-time $\tau$ should not be too large
  to control the finite difference error, and at the same time should
  not be too small, because otherwise the statistical error would be
  large, and the Markovianity would be disturbed.
\item Although the original dynamics defined by generator $\ml(t)$ is
  periodic, the reduced one defined by $\vect M(t)$ is not necessary
  periodic, i.e.~$\vect M(t) \neq \vect M(t+T)$ in general.  In the
  simulation of biomolecules, it is reasonable, in most cases to
  assume a time-scale separation in the system. For example, the
  conformational change of the molecules is much faster than the
  covalent bond vibrations. Therefore, if the discretion $\{ S_1,
  \cdots, S_n\}$ properly captures the slow dynamics of interest, in
  the sense that within the lag-time $\tau$, the system is fully
  relaxed to a local equilibrium in the set $S_i$, then it is clear
  that the reduce system is periodic. It suggests that the lag-time
  should be right in the spectrum gap: it is shorter than the dominant
  implied timescale, and longer than the fast-relaxing time-scales.
\end{enumerate}

@@

At this stage we do have any analytical result regarding the quality
of approximating Eqn.~\eqref{eqn:tmp1} by
Eqn.~\eqref{eqn:tmp2}.
It is therefore useful to investigate the discussed approximations
by a numerical example. The Markovianity of the reduce dynamics
was extensively discussed in e.g.~\cite{prinz2011markov}, so we 
will not touch this topic in the current paper. The choice of the lag-time
will be discussed later, and in this subsection we focus on the periodicity
of the reduced dynamics.
Throughout this paper, we take the alanine dipeptide system
under oscillatory EF, as an example, which was extensively studied by
Ref.~\cite{wang2014exploring}.
We choose the two dihedral angles $\phi$
and $\psi$ as slow variables (see Fig.~\ref{fig:tmp1}), and the discretization
is a uniform division of the $\phi$--$\psi$ plane. We denote the number of
division on each dihedral by $K$.

@@

% The definition of the conformations
% $\confaa$, $\confab$, $\confba$, $\confbb$ and $\confc$ are presented
% by Fig.~\ref{fig:tmp1}.  We are interested in the time-dependent
% probability of the conformations. In Fig.~\ref{fig:tmp2}, the brute
% force molecular dynamics simulation result
% (i.e.~Eqn.~\eqref{eqn:tmp1}) is compared with different discretization
% methods, which give different generator in Eqn.~\eqref{eqn:tmp2}. For
% clarity, only the probability of conformation $\confc$ is plotted.  In
% the figure, ``5 sets'' means the $\phi$-$\psi$ space is discretized by
% 5 sets as shown in Fig.~\ref{fig:tmp1}, ``Uni. $M=$'' stands for the
% uniform discretization on $\phi$-$\psi$ space, and $M$ denotes the
% number of bins on both $\phi$ and $\psi$ directions.  Both $M=10$ and
% 20 methods present very good accuracy.  Please notice that for $M=10$,
% for example, the number of discretized sets is $n=M^2=100$.  The
% lag-time in all cases is chosen to be $\tau = 0.5$~ps. A test with
% $\tau = 1.0$~ps shows good consistency with $\tau = 0.5$~ps.
% The generator
% $\vect M(t)$ is assumed to be periodic, and is estimated from the
% trajectories from $t=300$ to 1000~ps.

@@

The reduced dynamics will not provide a lot further interesting
information until we introduce the MSM to analyze the reduced dynamics. The main
difficulty is that the reduced dynamics is time-inhomogeneous, however, this can
be overcome by the Floquet theory.
The Floquet theory states that for a periodic system, e.g.~Eq.~\eqref{eqn:tmp2}, the solution can be written as
\begin{align}
  \label{eqn:tmp5}
  \vect p(t)  = \vect Q(t) e^{t\vect B} \vect p(0)
\end{align}
where $\vect Q(t)$ is periodic and $\vect B$ is a constant matrix.  To
analyze the spectrum, it is not difficult to infer the implied
timescales, on which the system relaxes to the periodic steady states
defined by $\vect Q(t)$.
Assume that the fundamental solution to Eqn.~\ref{eqn:tmp2} is denoted by $\boldsymbol \Phi(t)$, then
the matrix $\vect B$ can be expressed by
\begin{align}
  e^{T\vect B} = \boldsymbol \Phi^{-1}(0)\cdot \boldsymbol \Phi(T)
\end{align}
It is fully justified by choosing $ \boldsymbol \Phi(0) = I$, then
$e^{T\vect B}$ is estimated by evolving the fundamental solutions by
one period.
Since it is obvious that $Q(kT) = Q(0) = I,\ k\in\mathbb N$, if we only consider the state of the system
at time $kT$, then from Eq.~\eqref{eqn:tmp5} we have
\begin{align}
  \label{eqn:tmp5a}
  \vect p(kT)  = \boldsymbol\Phi^k(T) \,\vect p(0),
\end{align}
which is a time-homogeneous Markov process with transition matrix $\boldsymbol\Phi(T)$.
Please notice that in general this Markov process is not reversible.
Calculating the eigenvalues of $\boldsymbol\Phi(T)$ reveals the
implied timescales on which the non-equilibrium dynamics converges to
the periodic steady state.
The Eq.~\eqref{eqn:tmp5a} allows one to use the available techniques for time-homogeneous
Markov jump process (especially for irreversible processes)
to analyze the time-inhomogeneous dynamics Eq.~\eqref{eqn:tmp2}.
However, it can only resolve the processes that are longer than the period $T$.
Therefore, rather than going directly from Eq.~\eqref{eqn:tmp1} to Eq.~\eqref{eqn:tmp5a},
we would like to keep the intermediate step Eq.~\eqref{eqn:tmp2}.
In the following we compare the MSM with the brute force non-equilibrium simulation
by considering several properties that are of special interest.

@@

% At this stage we do have any analytical result regarding the quality
% of approximating Eqn.~\eqref{eqn:tmp1} by
% Eqn.~\eqref{eqn:tmp2}. However, it would be useful to do some
% numerical experiments. We have studied the alanine dipeptide system
% under oscillatory EF, which was studied by
% Ref.~\cite{wang2014exploring}. We use the two dihedral angles $\phi$
% and $\psi$ as slow variables. The discretization is the uniform division of
% the $\phi$--$\psi$ plan in to 2-dimentional bins. 
% The quality of the MSM~\eqref{eqn:tmp5a} can be checked by comparing the physical properties
% computed from the MSM and those computed by the brute force non-equilibrium molecular
% dynamics simulation from Ref.~\cite{wang2014exploring}.

>>>>

We consider  the following SDE form of the externally driven MD:
\begin{align}
  \label{eq:dis-1}
  d\vect x_t = \Big(-\nabla V(\vect x_t) + E(t) D(\vect x_t)\Big)dt + \sqrt{2\beta^{-1}} d\vect w_t, 
\end{align}
where the configurational space is denoted by $\Omega$, and the variable by
$\vect x$.   $V(\vect x)$ is the molecular interaction potential and
$E(t)D(x_t)$ the time-dependent external perturbation with the
$T$-periodic external field $E(t)$. $\beta$ is the inversed temperature,
i.e.~$\beta = 1/(k_B\mathcal T)$.
The propagation of probability
densities $\rho=\rho(\vect x,t)$ based on this kind of dynamics in the sense
of $\rho(\vect x,t)dx=\mathbb P[\vect x_t\in [\vect x,\vect x+d\vect x)]$ is governed by
Fokker-Planck equation:
\begin{align}
  \label{eq:dis-2}
  \frac{\partial \rho}{\partial t} = \ml^\dagger(t) \rho,
\end{align}
where $\ml^\dagger(t)$ is the adjoint of the generator
\begin{align}
  \label{eq:dis-3}
  \ml(t)=\beta^{-1}\Delta_{\vect x}+\Big(-\nabla V(\vect x) + E(t)D(\vect x)\Big)\cdot\nabla_{\vect x},
\end{align}
where $\Delta_{\vect x}$ denotes the Laplacian operator and $\nabla_{\vect x}$
the nabla-operator wrt to $\vect x$. 
The periodicity of the external driven indicates the periodicity of the generator,
i.e.~$\ml(t) = \ml(t+T)$.  Now
introduce a partition of the configurational space $\Omega$ into finite number of disjoint
sets $\{ \Omega_1, \cdots, \Omega_n\}$, which satisfy $\Omega = \cup_i \Omega_i$,
$\Omega_j\cap \Omega_j = \emptyset,\ \forall i\neq j$.
Now the original dynamics of the system can be discretized to a time-dependent
Markov jump process:
\begin{align}
  \label{eqn:tmp2}
  \frac{d\vect p(t)}{dt} = \vect M(t)\cdot \vect p(t)
\end{align}
where
\begin{align}
  \vect p(t) = (p_1(t), \cdots, p_n(t))^T, \quad \textrm{with}\ p_i(t) = \int_{\Omega_i} \rho(\vect x,t)d\vect x,
\end{align}
and the generator is approximated by finite difference scheme:
\begin{align}
  \label{eqn:tmp4}
  M_{ij}(t) \approx \frac{1}{\tau} \,[\, \mbp (\vect x_{t+\tau} \in \Omega_j \vert \vect x_{t} \in \Omega_i) - \delta_{ij} \,]
\end{align}
From Eqn.~\eqref{eq:dis-2} to Eqn.~\eqref{eqn:tmp2} involves several
approximations:
\begin{enumerate}
\item Eqn.~\eqref{eqn:tmp2} assumes Markovianity, which is not
  necessarily true for a reduce dynamics $\vect p(t)$, see e.g.~\cite{prinz2011markov}.
\item Numerical error from the finite difference introduced in
  Eqn.~\eqref{eqn:tmp4}. The lag-time $\tau$ should not be too large
  to control the finite difference error, and at the same time should
  not be too small, because otherwise the statistical error would be
  large, and the Markovianity would be disturbed.
\item Although the original dynamics defined by generator $\ml(t)$ is
  periodic, the reduced one defined by $\vect M(t)$ is not necessary
  periodic, i.e.~$\vect M(t) \neq \vect M(t+T)$ in general.  In the
  simulation of biomolecules, it is reasonable, in most cases to
  assume a time-scale separation in the system. For example, the
  conformational change of the molecules is much faster than the
  covalent bond vibrations. Therefore, if the discretion $\{ \Omega_1,
  \cdots, \Omega_n\}$ properly captures the slow dynamics of interest, in
  the sense that within the lag-time $\tau$, the system is fully
  relaxed to a local equilibrium in the set $\Omega_i$, then it is clear
  that the reduce system is periodic. It suggests that the lag-time
  should be right in the spectrum gap: it is shorter than the dominant
  implied timescale, and longer than the fast-relaxing time-scales.
\end{enumerate}

@@

At this stage we do have any analytical result regarding the quality
of approximating Eqn.~\eqref{eq:dis-2} by
Eqn.~\eqref{eqn:tmp2}.
It is therefore useful to investigate the discussed approximations
by a numerical example. The Markovianity of the reduce dynamics
was extensively discussed in e.g.~\cite{prinz2011markov}, so we 
will not touch this topic in the current paper. The choice of the lag-time
will be discussed later, and in this subsection we focus on the periodicity
of the reduced dynamics.
Throughout this paper, we take the alanine dipeptide system
under oscillatory EF, as an example, which was extensively studied by
Ref.~\cite{wang2014exploring}.
We choose the two dihedral angles $\phi$
and $\psi$ as slow variables (see Fig.~\ref{fig:tmp1}), and the discretization
is a uniform division of the $\phi$--$\psi$ plane. We denote the number of
division on each dihedral by $K$.

@@

% The definition of the conformations
% $\confaa$, $\confab$, $\confba$, $\confbb$ and $\confc$ are presented
% by Fig.~\ref{fig:tmp1}.  We are interested in the time-dependent
% probability of the conformations. In Fig.~\ref{fig:tmp2}, the brute
% force molecular dynamics simulation result
% (i.e.~Eqn.~\eqref{eq:dis-2}) is compared with different discretization
% methods, which give different generator in Eqn.~\eqref{eqn:tmp2}. For
% clarity, only the probability of conformation $\confc$ is plotted.  In
% the figure, ``5 sets'' means the $\phi$-$\psi$ space is discretized by
% 5 sets as shown in Fig.~\ref{fig:tmp1}, ``Uni. $M=$'' stands for the
% uniform discretization on $\phi$-$\psi$ space, and $M$ denotes the
% number of bins on both $\phi$ and $\psi$ directions.  Both $M=10$ and
% 20 methods present very good accuracy.  Please notice that for $M=10$,
% for example, the number of discretized sets is $n=M^2=100$.  The
% lag-time in all cases is chosen to be $\tau = 0.5$~ps. A test with
% $\tau = 1.0$~ps shows good consistency with $\tau = 0.5$~ps.
% The generator
% $\vect M(t)$ is assumed to be periodic, and is estimated from the
% trajectories from $t=300$ to 1000~ps.

@@

The reduced dynamics will not provide a lot further interesting
information until we introduce the MSM to analyze the reduced dynamics. The main
difficulty is that the reduced dynamics is time-inhomogeneous, however, this can
be overcome by the Floquet theory.
The Floquet theory states that for a periodic system, e.g.~Eq.~\eqref{eqn:tmp2}, the solution can be written as
\begin{align}
  \label{eqn:tmp5}
  \vect p(t)  = \vect Q(t) e^{t\vect B} \vect p(0)
\end{align}
where $\vect Q(t)$ is periodic and $\vect B$ is a constant matrix.  To
analyze the spectrum, it is not difficult to infer the implied
timescales, on which the system relaxes to the periodic steady states
defined by $\vect Q(t)$.
Assume that the fundamental solution to Eqn.~\ref{eqn:tmp2} is denoted by $\boldsymbol \Phi(t)$, then
the matrix $\vect B$ can be expressed by
\begin{align}
  e^{T\vect B} = \boldsymbol \Phi^{-1}(0)\cdot \boldsymbol \Phi(T)
\end{align}
It is fully justified by choosing $ \boldsymbol \Phi(0) = I$, then
$e^{T\vect B}$ is estimated by evolving the fundamental solutions by
one period.
Since it is obvious that $Q(kT) = Q(0) = I,\ k\in\mathbb N$, if we only consider the state of the system
at time $kT$, then from Eq.~\eqref{eqn:tmp5} we have
\begin{align}
  \label{eqn:tmp5a}
  \vect p(kT)  = \boldsymbol\Phi^k(T) \,\vect p(0),
\end{align}
which is a time-homogeneous Markov process with transition matrix $\boldsymbol\Phi(T)$.
Please notice that in general this Markov process is not reversible.
Calculating the eigenvalues of $\boldsymbol\Phi(T)$ reveals the
implied timescales on which the non-equilibrium dynamics converges to
the periodic steady state.
The Eq.~\eqref{eqn:tmp5a} allows one to use the available techniques for time-homogeneous
Markov jump process (especially for irreversible processes)
to analyze the time-inhomogeneous dynamics Eq.~\eqref{eqn:tmp2}.
However, it can only resolve the processes that are longer than the period $T$.
Therefore, rather than going directly from Eq.~\eqref{eq:dis-2} to Eq.~\eqref{eqn:tmp5a},
we would like to keep the intermediate step Eq.~\eqref{eqn:tmp2}.
In the following we compare the MSM with the brute force non-equilibrium simulation
by considering several properties that are of special interest.

@@

% At this stage we do have any analytical result regarding the quality
% of approximating Eqn.~\eqref{eq:dis-2} by
% Eqn.~\eqref{eqn:tmp2}. However, it would be useful to do some
% numerical experiments. We have studied the alanine dipeptide system
% under oscillatory EF, which was studied by
% Ref.~\cite{wang2014exploring}. We use the two dihedral angles $\phi$
% and $\psi$ as slow variables. The discretization is the uniform division of
% the $\phi$--$\psi$ plan in to 2-dimentional bins. 
% The quality of the MSM~\eqref{eqn:tmp5a} can be checked by comparing the physical properties
% computed from the MSM and those computed by the brute force non-equilibrium molecular
% dynamics simulation from Ref.~\cite{wang2014exploring}.
